---
title: "Portfolio" 
about:
  template: marquee  
  
execute:
  echo: false
---

## Project Overview

This project uses a Getis-Ord Gi* hotspot analysis to identify statistically significant clusters of median household income across the Seattle–Tacoma–Bellevue metropolitan area. Census tract-level income data from the American Community Survey (ACS) were used to examine patterns of spatial inequality and economic segregation within the region.

```{r setup, echo=FALSE}
#| message: false
#Install Packages

# install.packages("tidycensus")
# install.packages("tidyverse")
#install.packages("mapview")
# install.packages("tmap")
# install.packages("tigris")
# install.packages("crsuggest")
# install.packages("ggplot2")
# install.packages("spdep")
# install.packages("matrixStats")
# install.packages("rmapshaper")
# install.packages("SpatialAcc")

#Libraries

library("tidycensus")
library("tidyverse")
library(tigris)
options(tigris_use_cache = TRUE)
library(ggplot2)
library(mapview)
library(sf)
library(tmap)
library(crsuggest)
library(mapboxapi)
library(ggplot2)
library(spdep)
library(rmapshaper)
library(matrixStats)
library(SpatialAcc)

```

```{r,  echo=FALSE} 
#| message: false
#| warning: false 
seattle_cbsa <- core_based_statistical_areas(cb = TRUE, year = 2020) %>%
  filter(str_detect(NAME, "Seattle-Tacoma-Bellevue")) %>%
  st_transform(2285)   # Washington North (feet)

# Download Washington tracts with ACS median household income 
stb_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "WA",
  year = 2020,
  geometry = TRUE
) %>%
  st_transform(2285) %>%
  st_filter(seattle_cbsa, .predicate = st_within) %>%
  drop_na(estimate)

neighbors <- poly2nb(stb_tracts, queen = TRUE)

weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)

# Compute Getis-Ord Gi
localg_weights <- nb2listw(include.self(neighbors), zero.policy = TRUE)
stb_tracts$localG <- as.numeric(localG(stb_tracts$estimate, localg_weights))

# Classify hotspots
stb_tracts <- stb_tracts %>%
  mutate(hotspot = case_when(
    localG >= 2.56 ~ "High cluster",
    localG <= -2.56 ~ "Low cluster",
    TRUE ~ "Not significant"
  ))

# Map hotspots
ggplot(stb_tracts) +
  geom_sf(aes(fill = hotspot), color = "grey85", linewidth = 0.1) +
  scale_fill_manual(values = c(
    "High cluster" = "darkgreen",
    "Low cluster" = "goldenrod",
    "Not significant" = "grey90"
  )) +
  theme_void() +
  labs(
    title = "Getis-Ord Gi* Hotspot Analysis of Median Household Income",
    subtitle = "Seattle–Tacoma–Bellevue Metropolitan Area",
    fill = "Cluster Type",
    caption = "ACS 2016–2020 5-year estimates"
  )

```

### High-Income Clusters

High-income clusters are concentrated in the northeastern portion of the metropolitan area, particularly around Bellevue and Redmond. These hotspots indicate groups of census tracts with high median household incomes surrounded by neighboring tracts that also exhibit high incomes. The pattern reflects the concentration of technology and professional employment opportunities, strong educational systems, and some of the region's highest housing costs.

### Low-Income Clusters

Low-income clusters are located primarily south and west of Bellevue and extend toward Tacoma. These areas contain statistically significant concentrations of lower median household incomes. Contributing factors may include older housing stock, industrial land uses, lower average wages, and displacement pressures associated with rising housing costs in central Seattle.

## Interpretation

The hotspot analysis reveals a pronounced economic divide across the Seattle metropolitan region. Wealth is concentrated in the eastern portion of the study area, while lower-income communities are more prevalent in southern and western areas. These patterns suggest unequal access to economic opportunities and may reflect the long-term impacts of housing market dynamics, employment geography, and historical patterns of spatial inequality.

## Skills Demonstrated

- Spatial data analysis in R
- Census data acquisition using `tidycensus`
- Hotspot analysis using Getis-Ord Gi*
- GIS and spatial statistics
- Data visualization with `ggplot2` 
- Interpretation of socioeconomic spatial patterns

